### 2024-10-29 paper.tex
boilerplate and comments

\documentclass[11pt,sigconf]{iabart}
%\documentclass[10pt,sigconf,letterpaper]{acmart}
%\usepackage[vskip=1em,font=itshape,leftmargin=2em,rightmargin=2em]{quoting}
\usepackage[skip=4pt plus1pt]{parskip}
%\copyrightyear{2024}
%\acmYear{2024}
%\setcopyright{acmlicensed}\acmConference[IAB Network Management WS]{The IAB Next Era of Network Management Operations}{December 3, 2024}{}
%\acmBooktitle{The IAB next Era of Network Management Workshop}{December 3, 2024}{}

\begin{document}

\title{Managing IoT Devices with LwM2M}

\author{John Doe}
\email{john.doeericsson.com}
\orcid{0009-0008-2864-4269}
\affiliation{%
  \institution{Ericsson}
  \city{Jorvas}
  \country{Finland}
}

\begin{abstract}

% Common protocol themes: YANG, YANG-Push, NMDA, NETCONF, RESTCONF, CORECONF, SZTP, Call-Home, YANG-Library

% Current IETF network management topics:
% NETCONF (NETwork CONFiguration) WG
% NETCONF-next, RESTCONF-next, configuration of clients and servers, list pagination, transaction correlation, YANG-push transports (Protocols and YANG models)

% NETMOD (NETwork MODelling) WG
% YANG-next, YANG versioning, system configuration, data immutability (Language and YANG models)
% NMOP (Network Management Operations) WG
% YANG-push integration with Apache Kafka & time series databases, anomaly detection and incident management, digital map modelling

% Network Inventory (IVY) WG
% Hardware & software components, physical location, etc. correlating with existing IETF models (topolopy, service attachment points)

% Discussion topics will include, but are not limited to:
% 	•	Tooling, open source, experimentation, proof of concept, multi-vendor interoperability test (e.g., EANTC), and system integration
% 	•	Data consistency to support richer observability (Data & Knowledge)
% 	•	Integration issues with the business layer
% 	•	Automation, orchestration, and autonomy
% Goal is to Explore new requirements for future network management operations in a collaborative manner with the industry, network operators, and protocol engineers

% Check examples in slides: https://ripe89.ripe.net/wp-content/uploads/presentations/47-RIPE89_Next_Era_of_Network_Management_Operations_Benoit_Claise.pptx

In this paper we provide ...

\end{abstract}

\keywords{network management, operations}

\maketitle

\section{Introduction} \label{introduction}

The rapid evolution of network management protocols ...

\textbf{Current IETF protocols used}

% Perhaps mention ETAC and keep it light and only present what needs to be presented.
Our solution is built upon...

This paper references key protocols such as NMDA \cite{rfc8342}, NETCONF \cite{rfc6241}, and SNMP \cite{rfc3411} to explore advancements in network management.

\section{ETAC Overview?} \label{overview}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{figs/arch.pdf}
  \caption{General LwM2M Architecture}
  \label{fig:overall_architecture}
\end{figure}

% Sample placeholder image

This section presents an overview

\textbf{Participating Entities}

% high level
We define three primary entities

\begin{itemize}
  \item \textbf{one}: Typically a ...
\end{itemize}

\section{Protocol Insights} \label{insights}

\section{Protocol Evolution} \label{insights}

% Maybe GenAI and Intents?
% Maybe AI?
% RESTCONF/CORECONF (compression YANG/CBOR)

\section{Conclusions} \label{conclusions}

This paper provides an overview of ...

\section{Acknowledgments}

We would like to thank Ericsson for their support of this work. Special thanks to ...

\bibliographystyle{ACM-Reference-Format}
\bibliography{paper}

\end{document}
\endinput

### 2024-10-29 paper.tex
updated template

@@
 
-\title{Managing IoT Devices with LwM2M}
+\title{Network Management Experiences}
 
-\author{John Doe}
-\email{john.doeericsson.com}
+\author{J. Doe}
+\email{j.doe@ericsson.com}
+\orcid{0009-0008-2864-4269}
+\affiliation{%
+  \institution{Ericsson}
+  \city{Jorvas}
+  \country{Finland}
+}
+
+\author{J. Doe}
+\email{j.doe@ericsson.com}
+\orcid{0009-0008-2864-4269}
+\affiliation{%
+  \institution{Ericsson}
+  \city{Jorvas}
+  \country{Finland}
+}
+
+\author{J. Doe}
+\email{j.doe@ericsson.com}
 \orcid{0009-0008-2864-4269}
@@
 
-The rapid evolution of network management protocols ...
+The IAB workshop on the Next Era of Network Management Operations (NEMOPS) aims to foster dialogue between network operators and protocol developers, guiding the IETF in evolving network management protocols.
+This initiative seeks to evaluate past achievements and outline future requirements for network management operations.
 
-\textbf{Current IETF protocols used}
-
-% Perhaps mention ETAC and keep it light and only present what needs to be presented.
-Our solution is built upon...
-
-This paper references key protocols such as NMDA \cite{rfc8342}, NETCONF \cite{rfc6241}, and SNMP \cite{rfc3411} to explore advancements in network management.
+In this paper we introduce a cloud-native transport automation software solution designed to machine learning and open standards for advanced analytics and automation across microwave, IP, and optical fronthaul networks.
+The goal being to facilitate communication service providers the automation of their network operations, optimizing network performance, reducing costs, and enhancing end-user experiences.
 
 \section{ETAC Overview?} \label{overview}
+
+Depending on how deep we can get on what we can say at IETF, we may scope the paper accordingly
 
@@
 
+% Network Automation
+% Network Analytics/Visibility
+% Maybe ML?
+% RESTCONF/CORECONF (compression YANG/CBOR)
+
+\textbf{Current IETF protocols used}
+
+We use protocols such as NMDA \cite{rfc8342}, NETCONF \cite{rfc6241}, and SNMP \cite{rfc3411} to ...
+
+% Perhaps mention ETAC and keep it light and only present what needs to be presented.
+Our solution is built upon...
+
 \section{Protocol Evolution} \label{insights}
@@
 % Maybe GenAI and Intents?
-% Maybe AI?
-% RESTCONF/CORECONF (compression YANG/CBOR)
+% More standard interfaces?
+% More closed-loop Automation
+% Workflow Programmability and Network APIs (Ericsson message)
 

### 2024-11-04 paper.tex
initial version
restructuring based on meeting

--- 2024-10-29 paper.tex
@@
 
-\title{Managing IoT Devices with LwM2M}
+\title{TBD: NEMOPS Submission}
 
-\author{John Doe}
-\email{john.doeericsson.com}
+\author{J. Doe}
+\email{j.doe@ericsson.com}
+\orcid{0009-0008-2864-4269}
+\affiliation{%
+  \institution{Ericsson}
+  \city{Jorvas}
+  \country{Finland}
+}
+
+\author{J. Doe}
+\email{j.doe@ericsson.com}
+\orcid{0009-0008-2864-4269}
+\affiliation{%
+  \institution{Ericsson}
+  \city{Jorvas}
+  \country{Finland}
+}
+
+\author{J. Doe}
+\email{j.doe@ericsson.com}
 \orcid{0009-0008-2864-4269}
@@
 
+In this paper we provide ...
+
+\end{abstract}
+
+\keywords{network management, operations}
+
+\maketitle
+
+\section{Introduction} \label{introduction}
+
+The IAB workshop on the Next Era of Network Management Operations (NEMOPS) aims to foster communication between network operators and protocol developers, guiding the IETF in evolving network management protocols.
+This initiative seeks to evaluate past achievements and outline future requirements for network management operations.
+
+In this paper we introduce
+
+% a cloud-native transport automation software solution designed to machine learning and open standards for advanced analytics and automation across microwave, IP, and optical fronthaul networks.
+The goal being to facilitate communication service providers the automation of their network operations, optimizing network performance, reducing costs, and enhancing end-user experiences.
+
+\subsection{Challenges} \label{introduction}
+
+% This subsection will introduce the next one, giving a very brief overview.
+% Maybe we write this at the end
+
+\subsection{Overal Architecture} \label{overview}
+
+% Provide a high-level description of the standard components found in a network management controller or system responsible for managing network endpoints.
+Note: common cites.  \cite{rfc8342}, NETCONF \cite{rfc6241}, and SNMP \cite{rfc3411} to ...
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.5\textwidth]{figs/arch.pdf}
+  \caption{General Architecture}
+  \label{fig:overall_architecture}
+\end{figure}
+
+\section{Scalability} \label{scalability}
+
+% Challenges of scalability in network models/protocols/devices/systems
+% Optical equipment that can have up to 10,000 interfaces. Current models like YANG are inadequate due to limitations such as file databases without indexing.
+% There is a need for standardized network-level models that can be easily mapped to devices, facilitating scalability across various network architectures.
+% The integration of legacy systems remains a significant challenge, with many network nodes still relying on SNMP, necessitating adaptable and scalable solutions.
+
+\section{Telemetry} \label{telemetry}
+
+% The challenges of using UDP for telemetry.
+% Need for fine-grained control over data transmission.
+% Observability in retrieving records without overwhelming the DCN(?)
+% Handling of mmulti-protocol data and reporting mechanisms. Traps, YANG push, coap observe...
+% YANG to CBOR dictionary for compression and the implications of frequent notifications.
+% Signaling using lightweight protocols
+
+\section{Query Range} \label{queryrange}
+
+% Importance of protocol-agnostic network modeling, which allows for a broad query range.
+% Database storage and retrieval. Time-series storage. 
+% The meeting discussed the differences between UML-defined models and those used in YANG, focusing on how these models can operate in a protocol-neutral manner.
+
+\section{Network Management Evolution} \label{insights}
+
+% Maybe GenAI and Intents?
+% Other db/api query languages? GraphQL
+% Other, new standard interfaces?
+% More closed-loop Automation?
+% Workflow Programmability and Network APIs (Ericsson message)?
+
+\textbf{IETF and other SDOs}
+
+% ITUT difference between the models we define in UML and the ones we used in YANG, how things work in a protocol neutral way (MIB, YANG…) the industry was not ready for it back then.
+If you have a practical example …
+
+\section{Conclusions} \label{conclusions}
+
+This paper provides an overview of ...
+
+\section{Acknowledgments}
+
+We would like to thank Ericsson for their support of this work. Special thanks to ...
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{paper}
+
+\end{document}
+\endinput
+
+% NOTES
+
+% ------------------------------------------------------------------------
+% WS CALL
+% ------------------------------------------------------------------------
 % Common protocol themes: YANG, YANG-Push, NMDA, NETCONF, RESTCONF, CORECONF, SZTP, Call-Home, YANG-Library
@@
 
-In this paper we provide ...
+% ------------------------------------------------------------------------
 
-\end{abstract}
+% ------------------------------------------------------------------------
+% MEETING Nov4
+% ------------------------------------------------------------------------
 
-\keywords{network management, operations}
+% NEMOPS Paper
+% Signaling using lightweight protocols
+% IVY topic
+% Mention a bit the liaison, industry recognize center of competence. Connection between IETF, … ITUT
+% - Scalability
+% - Query range
+% - Telemetry
 
-\maketitle
+% ETAC Telemetry: UDP examples of struggles and issues, fine-grained granular control of what gets push.
+High level describe what we do currently.
+The reality of when developing these products you have to live in the legacy and the new world.
+Adaptors, many items in nodes are still SNMP, the reality is that u have to deal with a lot of legacy.
+Network modeling can be agnostic of the protocol used, network modeling can be abstracted to the method u use to the network elements.
 
-\section{Introduction} \label{introduction}
+% ITUT difference between the models we define in UML and the ones we used in YANG, how things work in a protocol neutral way (MIB, YANG…) the industry was not ready for it back then.
+If you have a practical example …
 
-The rapid evolution of network management protocols ...
+% Internally we have an adaptation layer towards the managed devices. 
+% Scott: standardize that?
+% When defining those network level models there should be a mapping between those network level models and device, so that they can be mapped easily.
+Scalability: optical equipment with 10k interfaces YANG does every poorly.
+File database with no indexing.
+Scott has a press about that.
+Redo the modeling of the bridge.
+% Compression: YANG to CBOR, something to think, if net elements notifications every 10s the smaller the better, fits also with the telemetry topic.
 
-\textbf{Current IETF protocols used}
-
-% Perhaps mention ETAC and keep it light and only present what needs to be presented.
-Our solution is built upon...
-
-This paper references key protocols such as NMDA \cite{rfc8342}, NETCONF \cite{rfc6241}, and SNMP \cite{rfc3411} to explore advancements in network management.
-
-\section{ETAC Overview?} \label{overview}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.5\textwidth]{figs/arch.pdf}
-  \caption{General LwM2M Architecture}
-  \label{fig:overall_architecture}
-\end{figure}
-
-% Sample placeholder image
-
-This section presents an overview
-
-\textbf{Participating Entities}
-
-% high level
-We define three primary entities
-
-\begin{itemize}
-  \item \textbf{one}: Typically a ...
-\end{itemize}
-
-\section{Protocol Insights} \label{insights}
-
-\section{Protocol Evolution} \label{insights}
-
-% Maybe GenAI and Intents?
-% Maybe AI?
-% RESTCONF/CORECONF (compression YANG/CBOR)
-
-\section{Conclusions} \label{conclusions}
-
-This paper provides an overview of ...
-
-\section{Acknowledgments}
-
-We would like to thank Ericsson for their support of this work. Special thanks to ...
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{paper}
-
-\end{document}
-\endinput
+% Telemetry: Observability to get back records.
+Do not want to kill the DCN.
+When you get stuff back, you have a value and put it somewhere, you can do traps, yang push, if you put it in a database, I don’t know if we have standard that maps it to that.
+Tool somewhere that maps.
+RESTCONF: Got stop time-series databases and relational databases.
+% GraphQL

### 2024-11-12 paper.tex
first major commit

@@
 
-\title{TBD: NEMOPS Submission}
+\title{Evolving Challenges and Solutions in Network Management}
 
-\author{J. Doe}
-\email{j.doe@ericsson.com}
+\author{Jaime Jiménez}
+\email{jaime.jimenez@ericsson.com}
 \orcid{0009-0008-2864-4269}
 \affiliation{%
-  \institution{Ericsson}
+  \institution{Ericsson, ER}
   \city{Jorvas}
@@
 
-\author{J. Doe}
-\email{j.doe@ericsson.com}
-\orcid{0009-0008-2864-4269}
+\author{Scott Mansfield}
+\email{scott.mansfield@ericsson.com}
 \affiliation{%
-  \institution{Ericsson}
+  \institution{Ericsson, BNEW TS ST}
   \city{Jorvas}
@@
 
-\author{J. Doe}
-\email{j.doe@ericsson.com}
-\orcid{0009-0008-2864-4269}
+\author{Raquel Rodriguez A}
+\email{raquel.a.rodriguez@ericsson.com}
 \affiliation{%
-  \institution{Ericsson}
+  \institution{Ericsson, ETAC}
+  \city{Jorvas}
+  \country{Finland}
+}
+
+\author{Mikko Pesonen}
+\email{mikko.pesonen@ericsson.com}
+\affiliation{%
+  \institution{Ericsson, ETAC}
+  \city{Jorvas}
+  \country{Finland}
+}
+
+\author{Vesa Torvinen}
+\email{Vesa.Torvinen@ericsson.com}
+\affiliation{%
+  \institution{Ericsson, ETAC}
   \city{Jorvas}
@@
 
-In this paper we provide ...
+This paper explores network management challenges and solutions, emphasizing scalability, telemetry, and security.
+It underscores the impact of AI and machine learning in enhancing operations via platforms like the Ericsson Transport Automation Controller (ETAC).
+The discussion includes legacy system complexities and diverse data sources, advocating for standardized schemas and efficient data streaming.
+It also covers the transition to zero-trust architectures and generative AI, highlighting the need for collaboration to create scalable, secure, and interoperable solutions.
+These insights aim to guide the industry in meeting modern telecommunications demands.
 
@@
 
-\keywords{network management, operations}
+\keywords{network management, scalability, telemetry, security, AI, zero-trust, interoperability}
 
@@
 
-The IAB workshop on the Next Era of Network Management Operations (NEMOPS) aims to foster communication between network operators and protocol developers, guiding the IETF in evolving network management protocols.
-This initiative seeks to evaluate past achievements and outline future requirements for network management operations.
+The IAB workshop on the Next Era of Network Management Operations (NEMOPS) serves as a pivotal platform for fostering dialogue between network operators and protocol developers.
+This initiative is instrumental in guiding the Internet Engineering Task Force (IETF) as it evolves network management protocols.
+The workshop's primary objectives are to assess past achievements and delineate future requirements for network management operations.
 
-In this paper we introduce
+In this paper, we introduce a comprehensive analysis of the current challenges and emerging solutions in network management.
+The subsequent sections delve into various facets of network management and operations.
+The \textbf{Overall Architecture} section \ref{overview} provides a detailed overview of the standard components within a network management controller, focusing on the Ericsson Transport Automation Controller (ETAC) and its integration with other network elements.
+The \textbf{Scalability} section \ref{scalability} examines the challenges of scaling network models and protocols, highlighting the need for standardized models.
+The \textbf{Telemetry} section \ref{telemetry} discusses the complexities of data transmission and the importance of efficient data streaming.
+The \textbf{Security} section \ref{security} addresses the diverse security protocols and the shift towards zero-trust architectures.
+Finally, the \textbf{Network Management Evolution} section \ref{insights} explores future trends, including the role of generative AI and new standard interfaces.
 
-% a cloud-native transport automation software solution designed to machine learning and open standards for advanced analytics and automation across microwave, IP, and optical fronthaul networks.
-The goal being to facilitate communication service providers the automation of their network operations, optimizing network performance, reducing costs, and enhancing end-user experiences.
+\subsection{ETAC overal Architecture} \label{overview}
 
-\subsection{Challenges} \label{introduction}
+Ericsson Transport Automation Controller (ETAC) ~\cite{ericsson-etac} is a cloud-native Transport Automation and SDN Controller that leverages artificial intelligence and machine learning to deliver advanced analytics and automation functionalities across microwave, IP, and optical fronthaul networks.
 
-% This subsection will introduce the next one, giving a very brief overview.
-% Maybe we write this at the end
-
-\subsection{Overal Architecture} \label{overview}
-
-% Provide a high-level description of the standard components found in a network management controller or system responsible for managing network endpoints.
-Note: common cites.  \cite{rfc8342}, NETCONF \cite{rfc6241}, and SNMP \cite{rfc3411} to ...
 \begin{figure}[h]
@@
 
+The \textbf{Ericsson Network Manager (ENM)} is a centralized platform responsible for managing network and service discovery, as well as collecting performance metrics.
+ENM interfaces with network elements like microwave nodes, known as \textbf{MINI-LINK}, to gather critical performance management (PM) data using \textbf{SFTP} \cite{RFC4253}.
+
+ENM also exposes network topology and service information to external systems through its Configuration Management Northbound Interface (CM NBI) over \textbf{HTTP/S} \cite{RFC7230}, enabling seamless integration with other platforms.
+This allows \textbf{Network Management Users} to monitor and control the network environment via a secure web-based interface.
+
+On the other side of the ecosystem lies the \textbf{ETAC} platform \cite{ericsson-etac}, an AI-enhanced tool designed to provide deeper insights and advanced analytics.
+ETAC communicates with ENM to retrieve PM data via \textbf{SFTP} \cite{RFC4253}, ensuring it has the necessary information to optimize network performance.
+ETAC also interfaces with MINI-LINK directly, using \textbf{SNMP} \cite{RFC1157} to receive real-time microwave link parameters.
+These parameters help ETAC maintain a granular view of the network's operational status.
+
+Furthermore, ETAC leverages \textbf{NETCONF/YANG} \cite{RFC6241, rfc6020} to configure and manage RDS (Radio Data System) and service parameters within MINI-LINK.
+This protocol enables precise, programmatic control over network devices, aligning with the industry's shift towards more automated and dynamic network configurations.
+
+Finally, both platforms provide secure \textbf{HTTP/S} \cite{RFC7230} interfaces for their respective user bases—ENM for Network Management Users and ETAC for ETAC-specific users.
+
 \section{Scalability} \label{scalability}
@@
 % The integration of legacy systems remains a significant challenge, with many network nodes still relying on SNMP, necessitating adaptable and scalable solutions.
+% Stateful connections vs stateless
+
+\textbf{YANG Schema Mount}
+
+Scalability in YANG presents significant challenges, as highlighted by Boyd \cite{boyd2023scalable}.
+The concern is about the viability of the current implementation of YANG Schema Mount \cite{RFC8528} and the need to support a mechanism that is better suited for partitioning the data in the hierarchical data tree.
+
+% This needs to be rewritten to human-level of understanding.
+Even an enhanced solution for this issue will really only be solved by a better underlying modeling implementation including ACID/transactional, time-series, object-relational views with a query-language and the ability to add triggers or embedded code that becomes part of the solution.
+The complexity of YANG is intensifying, largely driven by its hierarchical architecture.
+
+\textbf{Efficient Data Streaming for Analytics}
+
+The efficient and scalable transfer of analytics data from its source to post-processing systems remains a significant challenge.
+Legacy network elements predominantly rely on periodic data harvesting, which imposes unnecessary load on the network elements and delays data accessibility.
+To address this, data sources should implement active streaming of data to post-processing systems immediately upon production.
+This approach ensures that post-processing systems and closed-loop automation have access to data in near real-time.
 
@@
 
+% Concise Binary Object Representation (CBOR) compresses YANG modules for transmission via the Constrained Application Protocol (CoAP) \cite{RFC9254, rfc7252}.
+This is advantageous for devices in Low-Power and Lossy Networks (LLNs), which have limited power and memory.
+CBOR optimizes data transmission and enhances network efficiency in such environments.
+
+\textbf{Quality of data}
+
+Legacy network elements generate data that is not optimized for modern IT-style post-processing analytics systems.
+This data requires embedded metadata within its schema to support long-term storage and external post-processing, detailing the "what, where, when, and by whom," and identifying the precise source, such as a container, pod, application, node, or CaaS-cluster.
+
+Currently, the lack of alignment in analytics data schemas and metadata complicates and increases the cost of post-processing and analytics.
+Simple discrepancies, like varying timestamp formats, can cause significant issues.
+Ideally, standardized, extendable schemas and encodings for different analytics data types would simplify data processing across diverse vendor systems.
+
+Reliability of telemetry data may be less critical when used for visualization purposes and manual check-ups, however, it becomes crucial once telemetry is used as input for AI training and assisted network configuration
+
+\textbf{Diversity of data}
+
+Currently, analytic data is accessed through various APIs and follows diverse patterns (e.g., file harvesting, SNMP for FM data streaming, state notifications via NETCONF/RESTCONF, etc).
+This diversity complicates data collection and ingestion processes.
+There is a pressing need for an efficient, scalable, and universal transfer system to stream analytics data from its source to post-processing systems.
+This system must be optimized for bandwidth and CPU efficiency, ensure security, and be applicable to all types of analytics data streams, including metrics, events (e.g., alarms and notifications), logs, and state data.
+
+\textbf{Lineage of data}
+
+It is essential to establish the lineage and integrity of any record used for purposes beyond visualization.
+Utilizing data with uncertain integrity can introduce new attack vectors, such as enabling adversaries to exploit closed-loop automation by manipulating the input data.
+It is important to note that analytics data may flow directly from the source to the post-processing system; however, there are instances where data may traverse cloud provider PaaS functions.
+Therefore, a mechanism to ensure data integrity across multiple domains is necessary.
+
+\textbf{State-Data Handling}
+
+% This needs to be rewritten more clearly
+In passive data sources where analytics data is just exposed (as opposite to streamed) the data has to be actively harvested by external entities, thus there must be APIs or interfaces for collection of such data.
+In systems where data is streamed, the data sources (often involved in critical functions such as traffic handling) do not need to expose interfaces for harvesting analytics data, which makes the attack surface smaller and system more secure in general.
+
+Modern network elements have a lot of state-data which can not be efficiently leveraged through config model notifications.
+There are components and elements in the applications/data sources, which one does not to expose in a model, but still the analytics data about the state of such components should be available for post-processing systems.
+For example, one ideally would like to keep the model of cloud service realization agnostic, and use the same model for the cloud service, whether the application-instances of those services run in virtual machines or in container.
+Having to model compute resources in order to convey the state data would make it very tricky to separate realization and model from each other, and keep the model backwards compatible as the application realizations evolve and technologies change.
+
 \section{Query Range} \label{queryrange}
@@
 % The meeting discussed the differences between UML-defined models and those used in YANG, focusing on how these models can operate in a protocol-neutral manner.
+
+TBD!
+
+\section{Security} \label{security}
+
+\textbf{Security challenges}
+
+Security configuration in network management is complex due to the absence of a unified infrastructure.
+This complexity arises from the need to support multiple security protocols across diverse devices and vendors.
+Key points include:
+
+\begin{itemize}
+    \item \textbf{Diverse Protocols:} Network management applications must accommodate various security protocols, such as TLS, SSH, and username/password mechanisms.
+    \item \textbf{TLS Potential:} TLS, especially with client-server certificates, is a strong candidate for unified security but lacks universal support across all network management protocols and devices.
+%IS THIS TRUE???
+    \item \textbf{SSH Limitations:} there is no security infrastructure available that would distribute and help verifying the SSH public keys and facilitate flexible re-keying.The security configuration of SSH itself remains to be manual.
+    \item \textbf{Username/Password Usage:} Widely used for basic access and protocols like SNMPv3.
+Mechanisms for centralized authentication exists but cannot completely replace the need for local authentication because centralized mechanisms may not be always available.
+Manual security configuration leads easily to poor security practises as key renewals, password upgrades are easily neglected simply because it is error prone and expensive.
+
+\end{itemize}
+
+\textbf{Standardization process}
+
+The transition from standards to widespread network deployment is often very slow, particularly when new standards are to replace existing components.
+While implementing a new mechanism on a single device can be swift, updating all network devices is time-consuming.
+Management applications typically require market adoption or commitment to the new mechanism before implementation is deemed worthwhile.
+For this reason, phasing out legacy security mechanisms is challenging.
+
+\textbf{Zero-Trust Architecture}
+
+Ericsson develops products that are configured to use secure protocols and configurations by default.
+But we still see in legacy networks that sometimes the transport network is assumed to be secured, trusted, so that a lower degree of security is sometimes accepted.
+This may work in some networks but the trend is definitely towards zero-trust.
+
+Regulators and local legislation is also setting new requirements for critical infrastructure, such as telecom networks.
+The interest is not only on the security capabilities required from the network devices but also development processes, documentation, FOSS usage, security assurance and various aspects of the way how networks are deployed and managed.
+
+One unfortunate reality is that some network devices in life networks are actually very old.
+If the network operates correctly, any upgrade or change to it may start looking like a risk.
+This may lead to neglection of software and security upgrades.
+This could work in the walled-garden paradigm but definitely not anymore in zero-trust.
+Continuous deployments and upgrades are the only path to truly secure networks.
+And all this would need to be automated in secure way, covering the software upgrades, re-configurations and testing.
 
@@
 
-\textbf{IETF and other SDOs}
+\textbf{Generative AI}
 
-% ITUT difference between the models we define in UML and the ones we used in YANG, how things work in a protocol neutral way (MIB, YANG…) the industry was not ready for it back then.
-If you have a practical example …
+Generative AI (GenAI) is set to transform network management by incorporating AI systems that follow the autonomic networking principles outlined RFC7575.
+
+\begin{quote}
+"The fundamental goal is self-management, including self-configuration, self-optimization, self-healing, and self-protection." \cite{RFC7575}
+\end{quote}
+
+Agentic architectures, characterized by AI agents that autonomously discover and utilize tools provide dynamic interactions with complex systems, effectively adapting to changes that static codebases cannot.
+AI agents are capable of decomposing user intents into executable individual steps and can interact autonomously with other systems through well-known interfaces (e.g., Network and Management APIs).
+Agents can further federate and specialize in multi-agent systems, which improve on challenges such as hallucinations, specialization, and scalability.
+The interest in multi-agent systems was reflected during the past IETF 121 side meeting, as detailed in the ai4network agenda \cite{ai4network-agenda}.
+
+Agents are particularly useful in the telecommunications sector, where the complexity of specifications and codebases demands innovative solutions.
+Moreover, the current trend in 5G Networks is marked by a shift towards exposing network functionalities through APIs.
+This trend facilitates the integration of agentic systems that can dynamically interact with these APIs.
+
+\textbf{CoRECONF}
+
+The -CONF ecosystem comprises the following components:
+\begin{itemize}
+  \item \textbf{NETCONF} \cite{RFC6241}: Serializing YANG over a stateful TCP connection.    
+  \item \textbf{RESTCONF} \cite{RFC8040}: Serializing YANG over stateless HTTP.
+  \item \textbf{CoRECONF} \cite{draft-ietf-core-comi}: Serializing YANG modules in a CBOR \cite{RFC9254} map over stateless CoAP.
+\end{itemize}
+
+CoRECONF, in particular,is utilized by constrained devices in Low-Power and Lossy Networks, which are typically composed of numerous embedded devices with limited power and memory.
+Although primarily used for IoT, CoAP has also been specified for the signaling of DDoS-related telemetry \cite{RFC9132} and \cite{RFC9362}, as outlined in the now-concluded DOTS Working Group.
+
+This suggests potential applications for retrofitting CoAP on telemetry or management-type signaling within the network management domain.
+Specially in UDP-oriented evinronments as CoAP comes with an additional reliability mechanism and in environments where compression and smaller payloads are welcomed.
 
@@
 
-This paper provides an overview of ...
+This paper has explored the evolving landscape of network management, highlighting the challenges and solutions in scalability, telemetry, and security.
+The integration of advanced technologies such as AI and machine learning within platforms like the Ericsson Transport Automation Controller (ETAC) demonstrates the potential for enhanced analytics and automation in network operations.
+However, the complexity of legacy systems and the diversity of data sources present significant hurdles that require innovative approaches, such as standardized schemas and efficient data streaming mechanisms.
+
+The transition towards zero-trust architectures and the adoption of generative AI indicate a shift towards more secure and autonomous network management systems.
+These advancements necessitate a collaborative effort between industry stakeholders and standardization bodies to ensure seamless integration and widespread adoption.
+As the network management domain continues to evolve, the focus must remain on developing scalable, secure, and interoperable solutions that can adapt to the dynamic demands of modern telecommunications networks.
 
@@
 Do not want to kill the DCN.
-When you get stuff back, you have a value and put it somewhere, you can do traps, yang push, if you put it in a database, I don’t know if we have standard that maps it to that.
+When you get stuff back, you have a value and put it somewhere, you can do traps, yang push, if you put it in a database, I don't know if we have standard that maps it to that.
 Tool somewhere that maps.

### 2024-11-12 paper.tex
added agent image

@@
 
-\textbf{Generative AI}
+\textbf{AI Agents}
 
-Generative AI (GenAI) is set to transform network management by incorporating AI systems that follow the autonomic networking principles outlined RFC7575.
+Zero-shot LLMs are are already used to make operator insights more accessible, to streamline incident management or to generate task-specific code from natural language queries.
+As the industry progresses in the GenAI path, agentic architectures will be integrated into network AI systems, adhering to the autonomic networking principles specified in RFC7575.
 
@@
 
-Agentic architectures, characterized by AI agents that autonomously discover and utilize tools provide dynamic interactions with complex systems, effectively adapting to changes that static codebases cannot.
-AI agents are capable of decomposing user intents into executable individual steps and can interact autonomously with other systems through well-known interfaces (e.g., Network and Management APIs).
+Agentic systems are characterized by AI agents that are capable of decomposing user intents into executable individual steps and can interact autonomously with other systems through well-known interfaces (e.g., Network and Management APIs).
 Agents can further federate and specialize in multi-agent systems, which improve on challenges such as hallucinations, specialization, and scalability.
 The interest in multi-agent systems was reflected during the past IETF 121 side meeting, as detailed in the ai4network agenda \cite{ai4network-agenda}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.5\textwidth]{figs/agent.png}
+  \caption{Agentic System Architecture}
+  \label{fig:agent_architecture}
+\end{figure}
 

### 2024-11-12 paper.tex
eod commit

@@
 
-This paper explores network management challenges and solutions, emphasizing scalability, telemetry, and security.
-It underscores the impact of AI and machine learning in enhancing operations via platforms like the Ericsson Transport Automation Controller (ETAC).
-The discussion includes legacy system complexities and diverse data sources, advocating for standardized schemas and efficient data streaming.
-It also covers the transition to zero-trust architectures and generative AI, highlighting the need for collaboration to create scalable, secure, and interoperable solutions.
-These insights aim to guide the industry in meeting modern telecommunications demands.
+This paper presents a comprehensive analysis of the evolving challenges and emerging solutions in network management.
+We explore scalability issues in network models and protocols, emphasizing the limitations of current implementations like YANG and the need for standardized, extendable models.
+Telemetry complexities are discussed, focusing on data quality, diversity, and lineage, and the necessity for efficient data streaming and standardized schemas.
+Security challenges are addressed in the context of diverse protocols and the transition towards zero-trust architectures, highlighting the importance of unified security mechanisms and continuous updates.
+We also examine the future of network management, including the integration of generative AI and agentic architectures that adhere to autonomic networking principles, as well as new standard interfaces like CoRECONF.
+Our findings underscore the imperative for scalable, secure, and interoperable solutions that can adapt to the dynamic demands of modern telecommunications networks.
 
@@
 
-The IAB workshop on the Next Era of Network Management Operations (NEMOPS) serves as a pivotal platform for fostering dialogue between network operators and protocol developers.
-This initiative is instrumental in guiding the Internet Engineering Task Force (IETF) as it evolves network management protocols.
+The IAB workshop on the Next Era of Network Management Operations (NEMOPS) serves as a platform for discussion between network operators and protocol developers.
+This workshop is expected to guide the Internet Engineering Task Force (IETF) standards process.
 The workshop's primary objectives are to assess past achievements and delineate future requirements for network management operations.
 
-In this paper, we introduce a comprehensive analysis of the current challenges and emerging solutions in network management.
-The subsequent sections delve into various facets of network management and operations.
-The \textbf{Overall Architecture} section \ref{overview} provides a detailed overview of the standard components within a network management controller, focusing on the Ericsson Transport Automation Controller (ETAC) and its integration with other network elements.
+In this paper, we introduce a comprehensive analysis of the current challenges and emerging solutions in network management from the point of view of an SDN controller product.
+The subsequent sections delve into various aspects of network management and operations.
+The \textbf{Overall Architecture} section \ref{overview} provides a detailed overview of the standard components within a network management controller, the Ericsson Transport Automation Controller (ETAC).
 The \textbf{Scalability} section \ref{scalability} examines the challenges of scaling network models and protocols, highlighting the need for standardized models.
-The \textbf{Telemetry} section \ref{telemetry} discusses the complexities of data transmission and the importance of efficient data streaming.
-The \textbf{Security} section \ref{security} addresses the diverse security protocols and the shift towards zero-trust architectures.
-Finally, the \textbf{Network Management Evolution} section \ref{insights} explores future trends, including the role of generative AI and new standard interfaces.
+The \textbf{Telemetry} section \ref{telemetry} discusses the complexities of data transmission.
+The \textbf{Security} section \ref{security} raises some security challenges and explains the shift towards zero-trust architectures.
+Finally, the \textbf{Network Management Evolution} section \ref{insights} explores potential future trends, including the role of generative AI and new standard interfaces.
 
@@
 
-\section{Security} \label{security}
-
-\textbf{Security challenges}
+\section{Security challenges} \label{security}
 

### 2024-11-13 paper.tex
fixed references
v0.1

@@
-\documentclass[11pt,sigconf]{iabart}
+\documentclass[10pt,sigconf]{iabart}
 %\documentclass[10pt,sigconf,letterpaper]{acmart}
@@
 %\copyrightyear{2024}
-%\acmYear{2024}
-%\setcopyright{acmlicensed}\acmConference[IAB Network Management WS]{The IAB Next Era of Network Management Operations}{December 3, 2024}{}
+\acmYear{2024}
+\setcopyright{acmlicensed}\acmConference[IAB Network Management WS]{The IAB Next Era of Network Management Operations}{December 3, 2024}{}
 %\acmBooktitle{The IAB next Era of Network Management Workshop}{December 3, 2024}{}
@@
 
+\shortauthors{Jiménez et al}
+
 \author{Jaime Jiménez}
@@
 \orcid{0009-0008-2864-4269}
-\affiliation{%
+\affiliation{\small %
   \institution{Ericsson, ER}
@@
 \email{scott.mansfield@ericsson.com}
-\affiliation{%
+\affiliation{\small %
   \institution{Ericsson, BNEW TS ST}
@@
 \email{raquel.a.rodriguez@ericsson.com}
-\affiliation{%
+\affiliation{\small %
   \institution{Ericsson, ETAC}
@@
 \email{mikko.pesonen@ericsson.com}
-\affiliation{%
+\affiliation{\small %
   \institution{Ericsson, ETAC}
@@
 \email{Vesa.Torvinen@ericsson.com}
-\affiliation{%
+\affiliation{\small %
+  \institution{Ericsson, ETAC}
+  \city{Jorvas}
+  \country{Finland}
+}
+
+\author{Janne Karvonen}
+\email{janne.karvonen@ericsson.com}
+\affiliation{\small %
   \institution{Ericsson, ETAC}
@@
 
-\keywords{network management, scalability, telemetry, security, AI, zero-trust, interoperability}
+\keywords{Network Management, Scalability, Telemetry, Security, AI, Zero-Trust, Interoperability}
 
@@
 
-The \textbf{Ericsson Network Manager (ENM)} is a centralized platform responsible for managing network and service discovery, as well as collecting performance metrics.
-ENM interfaces with network elements like microwave nodes, known as \textbf{MINI-LINK}, to gather critical performance management (PM) data using \textbf{SFTP} \cite{RFC4253}.
+The \textbf{Ericsson Transport Automation Controller (ETAC)} \cite{ericsson-etac} adheres to the Open Transport SDN Reference Architecture as delineated by the TIP MUST project, aligning with the principles outlined in the Open Transport Architecture Whitepaper \cite{open_transport_architecture_whitepaper}.
+ETAC supports the roles of both the SDN Domain Controller and the SDN Hierarchical Controller, in accordance with the objectives of the Open Optical \& Packet Transport (OOPT) initiative \cite{oopt}.
+At high-level, it has two main Integration parts the Northound Integration part (NBI) and the Southbound Integration part (SBI) both of which use well-known standard interfaces and protocols like \textbf{NETCONF/YANG} \cite{RFC6241, rfc6020}, \textbf{RESTCONF} \cite{RFC8040}, \textbf{SNMP} \cite{RFC1157}, \textbf{SFTP} \cite{RFC4253} or \textbf{HTTP/S} \cite{RFC7230} towards APIs.
 
-ENM also exposes network topology and service information to external systems through its Configuration Management Northbound Interface (CM NBI) over \textbf{HTTP/S} \cite{RFC7230}, enabling seamless integration with other platforms.
-This allows \textbf{Network Management Users} to monitor and control the network environment via a secure web-based interface.
+On the SBI it communicates with the managed nodes and it also is capable of translating between specific device information models and the harmonized, standards-based information model used in a network database.
+The Network Intelligence layer builds on top of this harmonized model, implementing the analytics, automation and SDN control application supported in the system.
 
-On the other side of the ecosystem lies the \textbf{ETAC} platform \cite{ericsson-etac}, an AI-enhanced tool designed to provide deeper insights and advanced analytics.
-ETAC communicates with ENM to retrieve PM data via \textbf{SFTP} \cite{RFC4253}, ensuring it has the necessary information to optimize network performance.
-ETAC also interfaces with MINI-LINK directly, using \textbf{SNMP} \cite{RFC1157} to receive real-time microwave link parameters.
-These parameters help ETAC maintain a granular view of the network's operational status.
+On the NBI it offers an exposure layer to human users over a friendly GUI as well as to various OSS applications, providing access both to the network data and to the functionality supported in the system and enabling integration with other platforms.
 
-Furthermore, ETAC leverages \textbf{NETCONF/YANG} \cite{RFC6241, rfc6020} to configure and manage RDS (Radio Data System) and service parameters within MINI-LINK.
-This protocol enables precise, programmatic control over network devices, aligning with the industry's shift towards more automated and dynamic network configurations.
-
-Finally, both platforms provide secure \textbf{HTTP/S} \cite{RFC7230} interfaces for their respective user bases—ENM for Network Management Users and ETAC for ETAC-specific users.
+ETAC supports both SBI and NBI through implementations that accommodate legacy protocols and information models, whether defined by SOD or specific to vendors.
+It also provides real-time network observability, facilitating network analytics and closed-loop automation.
+It currently supports use cases that utilize its built-in real-time observability features to gain network insights through AI/ML and implement closed-loop automation in Transport Networks, all within a zero-trust framework.
 
@@
 
-% Challenges of scalability in network models/protocols/devices/systems
-% Optical equipment that can have up to 10,000 interfaces. Current models like YANG are inadequate due to limitations such as file databases without indexing.
-% There is a need for standardized network-level models that can be easily mapped to devices, facilitating scalability across various network architectures.
-% The integration of legacy systems remains a significant challenge, with many network nodes still relying on SNMP, necessitating adaptable and scalable solutions.
-% Stateful connections vs stateless
+% SCOTT please check this
+Scalability in network models, protocols, devices, and systems is a complex issue.
+For example, optical equipment can have up to 10,000 interfaces, but current models like YANG often fall short because they use file databases without indexing.
+This highlights the need for standardized network-level models that can be easily mapped to devices, improving scalability across different network architectures.
+Additionally, integrating legacy systems is a significant challenge, as many network nodes still rely on SNMP, which calls for adaptable and scalable solutions.
+The difference between stateful and stateless connections adds another layer of complexity to scalability efforts.
 
@@
 
-% This needs to be rewritten to human-level of understanding.
+% TODO: This needs to be rewritten more clearly
 Even an enhanced solution for this issue will really only be solved by a better underlying modeling implementation including ACID/transactional, time-series, object-relational views with a query-language and the ability to add triggers or embedded code that becomes part of the solution.
@@
 
+% TODO: BRIEF Intro
 % The challenges of using UDP for telemetry.
 % Need for fine-grained control over data transmission.
-% Observability in retrieving records without overwhelming the DCN(?)
-% Handling of mmulti-protocol data and reporting mechanisms. Traps, YANG push, coap observe...
+% Handling of mmulti-protocol data and reporting mechanisms. Traps, YANG push, webhooks
 % YANG to CBOR dictionary for compression and the implications of frequent notifications.
@@
 
-% Concise Binary Object Representation (CBOR) compresses YANG modules for transmission via the Constrained Application Protocol (CoAP) \cite{RFC9254, rfc7252}.
-This is advantageous for devices in Low-Power and Lossy Networks (LLNs), which have limited power and memory.
-CBOR optimizes data transmission and enhances network efficiency in such environments.
+% TODO: IMO the data part is to heavy with content specific for ETAC. We sould need some high-level general issues.
+% SCOTT, please input here.
 
@@
 
-% This needs to be rewritten more clearly
+% TODO: This needs to be rewritten more clearly
 In passive data sources where analytics data is just exposed (as opposite to streamed) the data has to be actively harvested by external entities, thus there must be APIs or interfaces for collection of such data.
@@
 
-\section{Query Range} \label{queryrange}
+% \section{Query Range} \label{queryrange}
 
-% Importance of protocol-agnostic network modeling, which allows for a broad query range.
-% Database storage and retrieval. Time-series storage. 
-% The meeting discussed the differences between UML-defined models and those used in YANG, focusing on how these models can operate in a protocol-neutral manner.
+% % Importance of protocol-agnostic network modeling, which allows for a broad query range.
+% % Database storage and retrieval. Time-series storage. 
+% % The meeting discussed the differences between UML-defined models and those used in YANG, focusing on how these models can operate in a protocol-neutral manner.
 
-TBD!
+% TBD!
+
+% TODO: Update Security by ETAC
 
@@
 This complexity arises from the need to support multiple security protocols across diverse devices and vendors.
-Key points include:
+Albeit rare, some challenges include:
 
@@
     \item \textbf{TLS Potential:} TLS, especially with client-server certificates, is a strong candidate for unified security but lacks universal support across all network management protocols and devices.
-%IS THIS TRUE???
-    \item \textbf{SSH Limitations:} there is no security infrastructure available that would distribute and help verifying the SSH public keys and facilitate flexible re-keying.The security configuration of SSH itself remains to be manual.
-    \item \textbf{Username/Password Usage:} Widely used for basic access and protocols like SNMPv3.
+% This was misswritten, my proposal, just say TLS/DTLS as universal stack (better no VPN)
+    \item \textbf{SSH Limitations:} There is no security infrastructure available that would distribute and help verifying the SSH public keys and facilitate flexible re-keying.The security configuration of SSH itself remains to be manual.
+% verify this statement
+    \item \textbf{Username/Password:} Widely used for basic access and protocols like SNMPv3.
 Mechanisms for centralized authentication exists but cannot completely replace the need for local authentication because centralized mechanisms may not be always available.
@@
 
-% Maybe GenAI and Intents?
-% Other db/api query languages? GraphQL
-% Other, new standard interfaces?
-% More closed-loop Automation?
-% Workflow Programmability and Network APIs (Ericsson message)?
+% TODO: Brief intro
 
@@
 
+% TODO: improve
 This paper has explored the evolving landscape of network management, highlighting the challenges and solutions in scalability, telemetry, and security.
@@
 
+% TODO: improve
 We would like to thank Ericsson for their support of this work. Special thanks to ...
@@
 \endinput
-
-% NOTES
-
-% ------------------------------------------------------------------------
-% WS CALL
-% ------------------------------------------------------------------------
-% Common protocol themes: YANG, YANG-Push, NMDA, NETCONF, RESTCONF, CORECONF, SZTP, Call-Home, YANG-Library
-
-% Current IETF network management topics:
-% NETCONF (NETwork CONFiguration) WG
-% NETCONF-next, RESTCONF-next, configuration of clients and servers, list pagination, transaction correlation, YANG-push transports (Protocols and YANG models)
-
-% NETMOD (NETwork MODelling) WG
-% YANG-next, YANG versioning, system configuration, data immutability (Language and YANG models)
-% NMOP (Network Management Operations) WG
-% YANG-push integration with Apache Kafka & time series databases, anomaly detection and incident management, digital map modelling
-
-% Network Inventory (IVY) WG
-% Hardware & software components, physical location, etc. correlating with existing IETF models (topolopy, service attachment points)
-
-% Discussion topics will include, but are not limited to:
-% 	•	Tooling, open source, experimentation, proof of concept, multi-vendor interoperability test (e.g., EANTC), and system integration
-% 	•	Data consistency to support richer observability (Data & Knowledge)
-% 	•	Integration issues with the business layer
-% 	•	Automation, orchestration, and autonomy
-% Goal is to Explore new requirements for future network management operations in a collaborative manner with the industry, network operators, and protocol engineers
-
-% Check examples in slides: https://ripe89.ripe.net/wp-content/uploads/presentations/47-RIPE89_Next_Era_of_Network_Management_Operations_Benoit_Claise.pptx
-
-% ------------------------------------------------------------------------
-
-% ------------------------------------------------------------------------
-% MEETING Nov4
-% ------------------------------------------------------------------------
-
-% NEMOPS Paper
-% Signaling using lightweight protocols
-% IVY topic
-% Mention a bit the liaison, industry recognize center of competence. Connection between IETF, … ITUT
-% - Scalability
-% - Query range
-% - Telemetry
-
-% ETAC Telemetry: UDP examples of struggles and issues, fine-grained granular control of what gets push.
-High level describe what we do currently.
-The reality of when developing these products you have to live in the legacy and the new world.
-Adaptors, many items in nodes are still SNMP, the reality is that u have to deal with a lot of legacy.
-Network modeling can be agnostic of the protocol used, network modeling can be abstracted to the method u use to the network elements.
-
-% ITUT difference between the models we define in UML and the ones we used in YANG, how things work in a protocol neutral way (MIB, YANG…) the industry was not ready for it back then.
-If you have a practical example …
-
-% Internally we have an adaptation layer towards the managed devices. 
-% Scott: standardize that?
-% When defining those network level models there should be a mapping between those network level models and device, so that they can be mapped easily.
-Scalability: optical equipment with 10k interfaces YANG does every poorly.
-File database with no indexing.
-Scott has a press about that.
-Redo the modeling of the bridge.
-% Compression: YANG to CBOR, something to think, if net elements notifications every 10s the smaller the better, fits also with the telemetry topic.
-
-% Telemetry: Observability to get back records.
-Do not want to kill the DCN.
-When you get stuff back, you have a value and put it somewhere, you can do traps, yang push, if you put it in a database, I don't know if we have standard that maps it to that.
-Tool somewhere that maps.
-RESTCONF: Got stop time-series databases and relational databases.
-% GraphQL

### 2024-11-13 paper.tex
first pass

@@
 \acmYear{2024}
-\setcopyright{acmlicensed}\acmConference[IAB Network Management WS]{The IAB Next Era of Network Management Operations}{December 3, 2024}{}
+%\setcopyright{acmlicensed}\acmConference[IAB Network Management WS]{The IAB Next Era of Network Management Operations}{December 3, 2024}{}
 %\acmBooktitle{The IAB next Era of Network Management Workshop}{December 3, 2024}{}
@@
 \title{Evolving Challenges and Solutions in Network Management}
-
+% Define short authors for page header
 \shortauthors{Jiménez et al}
@@
 
-This paper presents a comprehensive analysis of the evolving challenges and emerging solutions in network management.
-We explore scalability issues in network models and protocols, emphasizing the limitations of current implementations like YANG and the need for standardized, extendable models.
+This paper presents an analysis of the evolving challenges and emerging solutions in network management.
+We explore scalability issues in network models and protocols, emphasizing the some limitations of current implementations the need for standardized, extendable information models.
 Telemetry complexities are discussed, focusing on data quality, diversity, and lineage, and the necessity for efficient data streaming and standardized schemas.
 Security challenges are addressed in the context of diverse protocols and the transition towards zero-trust architectures, highlighting the importance of unified security mechanisms and continuous updates.
-We also examine the future of network management, including the integration of generative AI and agentic architectures that adhere to autonomic networking principles, as well as new standard interfaces like CoRECONF.
-Our findings underscore the imperative for scalable, secure, and interoperable solutions that can adapt to the dynamic demands of modern telecommunications networks.
+We also examine the future of network management, including the integration of generative AI and agentic architectures that adhere to autonomic networking principles, as well as the possibilities of new standard interfaces like CoRECONF.
+Our findings highlight the imperative for scalable, secure, and interoperable solutions that can adapt to the dynamic demands of modern telecommunications networks.
 
@@
 
-The IAB workshop on the Next Era of Network Management Operations (NEMOPS) serves as a platform for discussion between network operators and protocol developers.
+The IAB workshop on the Next Era of Network Management Operations (NEMOPS) serves as a platform for discussion between network operators, protocol experts and the general network management community.
 This workshop is expected to guide the Internet Engineering Task Force (IETF) standards process.
@@
 
-On the NBI it offers an exposure layer to human users over a friendly GUI as well as to various OSS applications, providing access both to the network data and to the functionality supported in the system and enabling integration with other platforms.
+On the NBI it offers an exposure layer to users over a friendly GUI as well as to various OSS applications, providing access both to the network data and to the functionality supported in the system and enabling integration with other platforms.
 
@@
 
-Legacy network elements generate data that is not optimized for modern IT-style post-processing analytics systems.
-This data requires embedded metadata within its schema to support long-term storage and external post-processing, detailing the "what, where, when, and by whom," and identifying the precise source, such as a container, pod, application, node, or CaaS-cluster.
+The alignment of analytics data schemas and metadata is essential for efficient post-processing and analytics.
+Minor differences, such as variations in timestamp formats, can pose challenges.
+The adoption of standardized, extendable schemas and encodings for various analytics data types would facilitate streamlined data processing across different vendor systems.
 
-Currently, the lack of alignment in analytics data schemas and metadata complicates and increases the cost of post-processing and analytics.
-Simple discrepancies, like varying timestamp formats, can cause significant issues.
-Ideally, standardized, extendable schemas and encodings for different analytics data types would simplify data processing across diverse vendor systems.
+Legacy network elements produce data that may not be fully compatible with modern IT-style post-processing analytics systems.
+To enhance compatibility, it is beneficial for this data to include embedded metadata providing details such as data originator.
 
-Reliability of telemetry data may be less critical when used for visualization purposes and manual check-ups, however, it becomes crucial once telemetry is used as input for AI training and assisted network configuration
+Lastly, while the reliability of telemetry data may be less critical for visualization and manual inspections, it becomes essential when utilized for AI training and assisted network configuration.
 
@@
 
-Currently, analytic data is accessed through various APIs and follows diverse patterns (e.g., file harvesting, SNMP for FM data streaming, state notifications via NETCONF/RESTCONF, etc).
-This diversity complicates data collection and ingestion processes.
-There is a pressing need for an efficient, scalable, and universal transfer system to stream analytics data from its source to post-processing systems.
-This system must be optimized for bandwidth and CPU efficiency, ensure security, and be applicable to all types of analytics data streams, including metrics, events (e.g., alarms and notifications), logs, and state data.
+Analytic data is often accessed through a variety of interfaces and standards.
+This variety presents challenges in streamlining data collection and ingestion processes.
+This is also true for including metrics which come in various forms like events, alarms, notifications, publications webhooks, logs, and other resource state data.
+
+There is therefore need for an efficient, scalable, and universal transfer system to stream analytics data from its source to post-processing systems.
 
@@
 Utilizing data with uncertain integrity can introduce new attack vectors, such as enabling adversaries to exploit closed-loop automation by manipulating the input data.
-It is important to note that analytics data may flow directly from the source to the post-processing system; however, there are instances where data may traverse cloud provider PaaS functions.
-Therefore, a mechanism to ensure data integrity across multiple domains is necessary.
+
+Ensuring data integrity across multiple domains is crucial, especially when analytics data traverses cloud provider PaaS functions before reaching post-processing systems.
 
@@
 \begin{itemize}
-    \item \textbf{Diverse Protocols:} Network management applications must accommodate various security protocols, such as TLS, SSH, and username/password mechanisms.
-    \item \textbf{TLS Potential:} TLS, especially with client-server certificates, is a strong candidate for unified security but lacks universal support across all network management protocols and devices.
-% This was misswritten, my proposal, just say TLS/DTLS as universal stack (better no VPN)
+    \item \textbf{Diverse Protocols:} Network management applications must accommodate various security protocols, such as TLS/DTLS, SSH, and username/password mechanisms.
+    \item \textbf{TLS/DTLS:} TLS/DTLS, particularly with client-server certificates, is a promising candidate for unified security.
+However, some vendors opt for alternative security techniques, such as VPNs.
     \item \textbf{SSH Limitations:} There is no security infrastructure available that would distribute and help verifying the SSH public keys and facilitate flexible re-keying.The security configuration of SSH itself remains to be manual.
-% verify this statement
-    \item \textbf{Username/Password:} Widely used for basic access and protocols like SNMPv3.
-Mechanisms for centralized authentication exists but cannot completely replace the need for local authentication because centralized mechanisms may not be always available.
-Manual security configuration leads easily to poor security practises as key renewals, password upgrades are easily neglected simply because it is error prone and expensive.
+% verify this statement there are things like hashicorp https://www.vaultproject.io
+    \item \textbf{Username/Password:} Commonly used for basic access and protocols like SNMPv3.
+Centralized authentication exists but cannot fully replace local authentication due to potential unavailability.
+Manual configuration often results in poor security practices, as key renewals and password updates are prone to errors and costly.
 
@@
 
-\textbf{Standardization process}
+\textbf{Legacy security}
 
 The transition from standards to widespread network deployment is often very slow, particularly when new standards are to replace existing components.
-While implementing a new mechanism on a single device can be swift, updating all network devices is time-consuming.
-Management applications typically require market adoption or commitment to the new mechanism before implementation is deemed worthwhile.
-For this reason, phasing out legacy security mechanisms is challenging.
+Additionally, management applications often necessitate market adoption or commitment to the new mechanism before its implementation is considered worthwhile.
+Consequently, phasing out legacy security mechanisms can be challenging.
 
@@
 
-% TODO: Brief intro
+The evolution of network management could be influenced by advancements in AI and machine learning, as well as potential retrofitting of other protocols intended for constrained environments.
+These developments promise enhanced automation and scalability for telecommunications networks.
 
@@
 Zero-shot LLMs are are already used to make operator insights more accessible, to streamline incident management or to generate task-specific code from natural language queries.
-As the industry progresses in the GenAI path, agentic architectures will be integrated into network AI systems, adhering to the autonomic networking principles specified in RFC7575.
+As the industry progresses in the recent GenAI path, agentic architectures will be integrated into network AI systems, adhering to the autonomic networking principles specified in RFC7575 \cite{RFC7575}.
 
 \begin{quote}
-"The fundamental goal is self-management, including self-configuration, self-optimization, self-healing, and self-protection." \cite{RFC7575}
+"The fundamental goal is self-management, including self-configuration, self-optimization, self-healing, and self-protection." 
 \end{quote}
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figs/agent.png}
-  \caption{Agentic System Architecture}
+  \includegraphics[width=0.5\textwidth]{figs/agent.pdf}
+  \caption{Agent Architecture}
   \label{fig:agent_architecture}
@@
 
-The -CONF ecosystem comprises the following components:
+CoRECONF, in particular,is utilized by constrained devices in Low-Power and Lossy Networks, which are typically composed of numerous embedded devices with limited power and memory.
+
+The the Constrained Application Protocol (CoAP) \cite{RFC7252}, a component of CoRECONF, is primarily utilized in IoT applications, it has also been specified for signaling DDoS-related telemetry \cite{RFC9132} and \cite{RFC9362}, as documented by the now-concluded DOTS Working Group.
+
+Within the network management community, the -CONF ecosystem is predominantly characterized by the widespread deployment of NETCONF and RESTCONF for network management tasks.
+In contrast, CoRECONF is not as widely know.
+The main differences being the application protocol and the serializations:
+
 \begin{itemize}
@@
 \end{itemize}
-
-CoRECONF, in particular,is utilized by constrained devices in Low-Power and Lossy Networks, which are typically composed of numerous embedded devices with limited power and memory.
-Although primarily used for IoT, CoAP has also been specified for the signaling of DDoS-related telemetry \cite{RFC9132} and \cite{RFC9362}, as outlined in the now-concluded DOTS Working Group.
 
@@
 % TODO: improve
+
 This paper has explored the evolving landscape of network management, highlighting the challenges and solutions in scalability, telemetry, and security.

### 2024-11-14 paper.tex

@@
 Legacy network elements predominantly rely on periodic data harvesting, which imposes unnecessary load on the network elements and delays data accessibility.
-To address this, data sources should implement active streaming of data to post-processing systems immediately upon production.
+To address this, the data sources should implement active streaming of analytics data to post-processing systems immediately upon production.
 This approach ensures that post-processing systems and closed-loop automation have access to data in near real-time.
@@
 The alignment of analytics data schemas and metadata is essential for efficient post-processing and analytics.
-Minor differences, such as variations in timestamp formats, can pose challenges.
-The adoption of standardized, extendable schemas and encodings for various analytics data types would facilitate streamlined data processing across different vendor systems.
+Minor differences, such as variations in timestamp formats, can pose challenges in post-processing systems when correlating data from different sources.
+The adoption of standardized, extendable schemas and encodings for various analytics data types and streams would facilitate streamlined data processing across different vendor systems.
 
-Legacy network elements produce data that may not be fully compatible with modern IT-style post-processing analytics systems.
-To enhance compatibility, it is beneficial for this data to include embedded metadata providing details such as data originator.
-
-Lastly, while the reliability of telemetry data may be less critical for visualization and manual inspections, it becomes essential when utilized for AI training and assisted network configuration.
-
+Legacy network elements produce analytics data that may not be fully compatible with modern IT-style post-processing analytics systems where post-processing takes place outside the network element.
+The structure of data records of analytics data streams should be enhanced to include enough embedded metadata to ensure that data can be stored and post-processed in external systems.
 \textbf{Diversity of data}
@@
 
-There is therefore need for an efficient, scalable, and universal transfer system to stream analytics data from its source to post-processing systems.
+There is therefore need for an efficient, scalable, and universal transfer method of streaming theanalytics data from data sources to post-processing systems and other consumers.
 
@@
 
-It is essential to establish the lineage and integrity of any record used for purposes beyond visualization.
+It should also be possible to establish the lineage and integrity of any data to ensure that it can be used safely for any purpose beyond simple visualization such as closed loop automation, AI-training and input for various automatic decision making processes.
+
 Utilizing data with uncertain integrity can introduce new attack vectors, such as enabling adversaries to exploit closed-loop automation by manipulating the input data.

### 2024-11-14 paper.tex
jari's review

--- 2024-11-13 paper.tex
@@
 \title{Evolving Challenges and Solutions in Network Management}
-% Define short authors for page header
+
 \shortauthors{Jiménez et al}
@@
 
-% SCOTT please check this
-Scalability in network models, protocols, devices, and systems is a complex issue.
-For example, optical equipment can have up to 10,000 interfaces, but current models like YANG often fall short because they use file databases without indexing.
-This highlights the need for standardized network-level models that can be easily mapped to devices, improving scalability across different network architectures.
-Additionally, integrating legacy systems is a significant challenge, as many network nodes still rely on SNMP, which calls for adaptable and scalable solutions.
-The difference between stateful and stateless connections adds another layer of complexity to scalability efforts.
+\textbf{YANG Scalability}
 
-\textbf{YANG Schema Mount}
+Scalability in YANG is challenging, as noted by Boyd \cite{boyd2023scalable}.
+YANG validation can be computationally intensive, and performance degrades with large data stores.
+Models must be tailored for large-scale devices.
 
-Scalability in YANG presents significant challenges, as highlighted by Boyd \cite{boyd2023scalable}.
-The concern is about the viability of the current implementation of YANG Schema Mount \cite{RFC8528} and the need to support a mechanism that is better suited for partitioning the data in the hierarchical data tree.
+The complexity of YANG is largely driven by its hierarchical architecture.
+For example, optical equipment can have tens of thousands of interfaces, but current models like YANG often fall short because they use file databases without indexing.
+There is also concern about the viability of the current implementation of YANG Schema-mount \cite{RFC8528} for Optical Network Units (ONU) YANG models, as it seems incompatible with the ONU template technique without editing the YANG modules.
 
-% TODO: This needs to be rewritten more clearly
-Even an enhanced solution for this issue will really only be solved by a better underlying modeling implementation including ACID/transactional, time-series, object-relational views with a query-language and the ability to add triggers or embedded code that becomes part of the solution.
-The complexity of YANG is intensifying, largely driven by its hierarchical architecture.
+Some modeling challenges can be mitigated by altering the model's structure; however, this often results in backward incompatibility.
+While YANG features can be adapted or extended to address these issues, doing so introduces additional complexity and also leads to backward incompatibility.
 
@@
 
-The efficient and scalable transfer of analytics data from its source to post-processing systems remains a significant challenge.
 Legacy network elements predominantly rely on periodic data harvesting, which imposes unnecessary load on the network elements and delays data accessibility.
-To address this, data sources should implement active streaming of data to post-processing systems immediately upon production.
-This approach ensures that post-processing systems and closed-loop automation have access to data in near real-time.
+To address this, data sources should implement active streaming of data to post-processing systems immediately upon production, this will also ensure that closed-loop automation systems have access to data in near real-time.
 
@@
 
-% \section{Query Range} \label{queryrange}
-
-% % Importance of protocol-agnostic network modeling, which allows for a broad query range.
-% % Database storage and retrieval. Time-series storage. 
-% % The meeting discussed the differences between UML-defined models and those used in YANG, focusing on how these models can operate in a protocol-neutral manner.
-
-% TBD!
-
 % TODO: Update Security by ETAC
@@
 
+Zero trust security, or zero trust architecture (ZTA), mandates "never trust, always verify," ensuring no default trust for users or devices, even on permissioned networks.
 Ericsson develops products that are configured to use secure protocols and configurations by default.
-But we still see in legacy networks that sometimes the transport network is assumed to be secured, trusted, so that a lower degree of security is sometimes accepted.
-This may work in some networks but the trend is definitely towards zero-trust.
 
-Regulators and local legislation is also setting new requirements for critical infrastructure, such as telecom networks.
-The interest is not only on the security capabilities required from the network devices but also development processes, documentation, FOSS usage, security assurance and various aspects of the way how networks are deployed and managed.
-
-One unfortunate reality is that some network devices in life networks are actually very old.
-If the network operates correctly, any upgrade or change to it may start looking like a risk.
-This may lead to neglection of software and security upgrades.
-This could work in the walled-garden paradigm but definitely not anymore in zero-trust.
-Continuous deployments and upgrades are the only path to truly secure networks.
-And all this would need to be automated in secure way, covering the software upgrades, re-configurations and testing.
+Some operational networks still rely on outdated devices, and any attempt to upgrade or modify these systems is often perceived as a risk, resulting in the neglect of essential software and security updates.
+While this approach might have been viable within a walled-garden security paradigm, it is incompatible with the zero-trust framework.
 
@@
 
-Zero-shot LLMs are are already used to make operator insights more accessible, to streamline incident management or to generate task-specific code from natural language queries.
-As the industry progresses in the recent GenAI path, agentic architectures will be integrated into network AI systems, adhering to the autonomic networking principles specified in RFC7575 \cite{RFC7575}.
+The advent of Large Language Models (LLMs) presents an opportunity for autonomous networking and network management.
+LLMs are already employed to enhance operator insights, streamline incident management, and generate task-specific code from natural language queries.
+As the industry advances along the GenAI path, these capabilities will be further augmented by integrating agentic architectures into network AI systems.
+
+LLMs enable the creation of AI agents that are capable of decomposing user intents into executable individual steps and can interact autonomously with other systems through well-known interfaces (e.g., Network and Management APIs).
+This integration aligns with the autonomic networking principles outlined in RFC7575 \cite{RFC7575}, facilitating more efficient and adaptive network management.
 
@@
 
-Agentic systems are characterized by AI agents that are capable of decomposing user intents into executable individual steps and can interact autonomously with other systems through well-known interfaces (e.g., Network and Management APIs).
-Agents can further federate and specialize in multi-agent systems, which improve on challenges such as hallucinations, specialization, and scalability.
+Furthermore, agents can federate and specialize in multi-agent systems, which improve on challenges such as hallucinations, specialization, and scalability.
 The interest in multi-agent systems was reflected during the past IETF 121 side meeting, as detailed in the ai4network agenda \cite{ai4network-agenda}.
@@
 
+Challenges in this domain include aligning intents from various sources that may have conflicting objectives and ensuring AI outputs are free from hallucinations.
+The latter is addressed by incorporating fact-checking, dataset tagging, human-in-the-loop solutions, and other verification mechanisms.
+On top of that it is likely that there will be incremental improvements on each of the components of an agent, both in planning, memory and tooling.
+
+Moreover, there are limitations in terms of control of the most advanced LLMs which are the foundation of these types of agents, it is likely that privacy concerns and lack of in-house LLM deployments limits the usage.
+
 \textbf{CoRECONF}
 
-CoRECONF, in particular,is utilized by constrained devices in Low-Power and Lossy Networks, which are typically composed of numerous embedded devices with limited power and memory.
+There is potential for retrofitting Internet of Things (IoT) oriented protocols on telemetry or management-type signaling within the network management domain.
+For example in UDP evinronments and in environments where compression and smaller payloads are welcomed.
+
+In particular, CoRECONF is utilized by constrained devices in Low-Power and Lossy Networks, which are typically composed of numerous embedded devices with limited power and memory.
 
@@
 \end{itemize}
-
-This suggests potential applications for retrofitting CoAP on telemetry or management-type signaling within the network management domain.
-Specially in UDP-oriented evinronments as CoAP comes with an additional reliability mechanism and in environments where compression and smaller payloads are welcomed.
 
@@
 
-% TODO: improve
-We would like to thank Ericsson for their support of this work. Special thanks to ...
+We would like to express our gratitude to Ericsson for their support of this work. Special thanks to Jari Arkko for his insightful review.
 

### 2024-11-14 paper.tex
ETAC edits

@@
 
-\subsection{ETAC overal Architecture} \label{overview}
+\subsection{ETAC overall Architecture} \label{overview}
 
@@
 
-ETAC supports both SBI and NBI through implementations that accommodate legacy protocols and information models, whether defined by SOD or specific to vendors.
+ETAC supports both SBI and NBI through implementations that accommodate legacy protocols and information models, whether defined by a Standards Developing Organization (SDO) or by specific to vendors.
 It also provides real-time network observability, facilitating network analytics and closed-loop automation.
@@
 
-% TODO: BRIEF Intro
-% The challenges of using UDP for telemetry.
-% Need for fine-grained control over data transmission.
-% Handling of mmulti-protocol data and reporting mechanisms. Traps, YANG push, webhooks
-% YANG to CBOR dictionary for compression and the implications of frequent notifications.
-% Signaling using lightweight protocols
+\textbf{Protocol efficiency}
 
-% TODO: IMO the data part is to heavy with content specific for ETAC. We sould need some high-level general issues.
-% SCOTT, please input here.
+In high-rate telemetry use cases, line protocol efficiency is a key consideration.
+With NETCONF, the verbose nature of the XML encoding implies both added processing overhead in server and client, and increased bandwidth requirement on the DCN channel.
+Techniques for reducing the processing and bandwidth requirements should be considered in future work.
+One example of the importance of efficiency is is gRPC, which has gained popularity in telemetry applications largely due to its line protocol efficiency.
 
-\textbf{Quality of data}
+Similarly, the choice of transport protocol has a key impact on the processing and bandwidth requirements.
+TCP, even though more reliable, is heavier in processing and on the line so often UDP would be preferred for transport in telemetry applications.
+With UDP, however, additional mechanisms may be needed on application layer to compensate for the unreliable transport and to handle e.g.
+packet drops and reorderings, in order not to lose critical data, and these mechanism may in turn result in losing the benefits of the lighter transport layer.
+There's thus a fine balance between when to use UDP vs.
+TCP and the choice should be based on careful consideration of all these factors.
 
-The alignment of analytics data schemas and metadata is essential for efficient post-processing and analytics.
-Minor differences, such as variations in timestamp formats, can pose challenges.
-The adoption of standardized, extendable schemas and encodings for various analytics data types would facilitate streamlined data processing across different vendor systems.
+\textbf{Quality and integrity of the data}
 
-Legacy network elements produce data that may not be fully compatible with modern IT-style post-processing analytics systems.
-To enhance compatibility, it is beneficial for this data to include embedded metadata providing details such as data originator.
+The alignment of telemetry data across different sources and subsystems is essential for efficient post-processing of the data.
+This calls for more coordination in the development of YANG data models, e.g.
+in order to be able to represent the necessary associations between YANG modules developed in different groups and the schemas of the records, so that the data can be linked correctly and efficiently in post-processing.
+Also what should be taken into account is the lineage and integrity in the streamed data records, e.g.
+to not introduce new attack vectors by enabling adversaries to exploit closed-loop automation by manipulating the telemetry data.
 
-Lastly, while the reliability of telemetry data may be less critical for visualization and manual inspections, it becomes essential when utilized for AI training and assisted network configuration.
+\textbf{Flexibility}
 
-\textbf{Diversity of data}
+The amount of configuration, state and performance data available and produced in modern network elements is ever-increasing, and ability to have all this data available for network management, analytics and automation applications is crucial in managing the complexity of networks today and in the future.
+Thus, the protocols and mechanisms developed for streaming telemetry should be optimized to support high-rate streaming of large volumes of data efficiently.
 
-Analytic data is often accessed through a variety of interfaces and standards.
-This variety presents challenges in streamlining data collection and ingestion processes.
-This is also true for including metrics which come in various forms like events, alarms, notifications, publications webhooks, logs, and other resource state data.
-
-There is therefore need for an efficient, scalable, and universal transfer system to stream analytics data from its source to post-processing systems.
-
-\textbf{Lineage of data}
-
-It is essential to establish the lineage and integrity of any record used for purposes beyond visualization.
-Utilizing data with uncertain integrity can introduce new attack vectors, such as enabling adversaries to exploit closed-loop automation by manipulating the input data.
-
-Ensuring data integrity across multiple domains is crucial, especially when analytics data traverses cloud provider PaaS functions before reaching post-processing systems.
-
-\textbf{State-Data Handling}
-
-% TODO: This needs to be rewritten more clearly
-In passive data sources where analytics data is just exposed (as opposite to streamed) the data has to be actively harvested by external entities, thus there must be APIs or interfaces for collection of such data.
-In systems where data is streamed, the data sources (often involved in critical functions such as traffic handling) do not need to expose interfaces for harvesting analytics data, which makes the attack surface smaller and system more secure in general.
-
-Modern network elements have a lot of state-data which can not be efficiently leveraged through config model notifications.
-There are components and elements in the applications/data sources, which one does not to expose in a model, but still the analytics data about the state of such components should be available for post-processing systems.
-For example, one ideally would like to keep the model of cloud service realization agnostic, and use the same model for the cloud service, whether the application-instances of those services run in virtual machines or in container.
-Having to model compute resources in order to convey the state data would make it very tricky to separate realization and model from each other, and keep the model backwards compatible as the application realizations evolve and technologies change.
-
-% TODO: Update Security by ETAC
+The potentially high volume of data also calls for ability to have fine-grain control on what data gets streamed.
+Different applications are interested in different subsets of the data, may need it at different frequency or their needs may e.g.
+be adaptive according to the condition of the network.
+Also, there's often a tradeoff between the needs of an application and what can be delivered e.g.
+due to the limited DCN bandwidth available in the network.
+Being able to exactly
+choose what gets streamed and when helps to overcome these challenges.
 
@@
 However, some vendors opt for alternative security techniques, such as VPNs.
-    \item \textbf{SSH Limitations:} There is no security infrastructure available that would distribute and help verifying the SSH public keys and facilitate flexible re-keying.The security configuration of SSH itself remains to be manual.
-% verify this statement there are things like hashicorp https://www.vaultproject.io
+    \item \textbf{SSH Limitations:} Security infrastructure to distribute and help verifying the SSH public keys and to facilitate flexible re-keying is not commonly in use, probably due to complexity.The security configuration of SSH itself remains to be manual.
     \item \textbf{Username/Password:} Commonly used for basic access and protocols like SNMPv3.
@@
 
-Some operational networks still rely on outdated devices, and any attempt to upgrade or modify these systems is often perceived as a risk, resulting in the neglect of essential software and security updates.
-While this approach might have been viable within a walled-garden security paradigm, it is incompatible with the zero-trust framework.
+In legacy networks, sometimes the transport network is assumed to be secured, trusted, so that lower security can be accepted.
+This may work in some networks but the trend is definitely towards zero-trust.
+Regulators and local legislation is also setting new requirements for critical infrastructure, such as telecom networks.
+The interest is not only on the security capabilities required from the network devices but also development processes, documentation, FOSS usage, security assurance and various aspects of the way how networks are deployed and managed.
 
@@
 
-The evolution of network management could be influenced by advancements in AI and machine learning, as well as potential retrofitting of other protocols intended for constrained environments.
+The evolution of network management and automation could be influenced by advancements in AI and machine learning, as well as potential retrofitting of other protocols intended for constrained environments.
 These developments promise enhanced automation and scalability for telecommunications networks.
+
+\textbf{AI/ML/Analytics for enhanced observability and network insights}
+
+Legacy network management platforms ingest different types of PM, FM, events data generated by network elements and in most cases legacy network managers leave the interpretation and correlation of that data to humans.
+
+AI/ML techniques like anomaly detection, classification, prediction, etc can be run over time series data, network topology and inventory data to detect and predict network events by looking at anomalies, patterns in data, forecasting trend, etc.
+
+The AI/MLs/Analytics generated data can then be used to produce network insights in general, being some of those also actionable (closed loop automations).
 
@@
 This trend facilitates the integration of agentic systems that can dynamically interact with these APIs.
+
+The multi-agent domain could also have applications in networks, CSP Transport Networks are typically very heterogeneous with a combination of Transport domains (Microwave, IP, Optical) and equipment vendor.
+Therefore domain/vendor specific agents can be trained.
+SDN platforms can then perform orchestration the specialized agents to achieve overall better performance in heterogenous networks.
 
